|

### 2024-01-07 paper.Rmd
reduce to 32 + 3 references (almost there)

\textcolor{red}{Diagnosing a model is an important part of building an appropriate model.} In linear regression analysis, studying the residuals from a model fit is a common diagnostic activity. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

@@

\textcolor{red}{Scatterplots of residuals against the fitted values, and each of the explanatory variables, are commonly used to scrutinize their relationships}. If there are any visually discoverable associations, the model is potentially inadequate or incorrectly specified. \textcolor{red}{We can also potentially discover patterns not directly connected to a linear model assumption from these residual plots, such as the discreteness or skewness of the fitted values, and outliers.} To read residual plots, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. Statistical tests were developed to provide objective assessment, for example, of non-linear patterns [e.g. @ramsey1969tests], and heteroskedasticity  [e.g. @breusch1979simple].

@@

\textcolor{red}{Additionally, relying solely on the subjective assessment of a single plot can be problematic. People will almost always see a pattern \citep[see][]{kahneman2011thinking}, so the question that really needs answering is whether any pattern perceived is consistent with randomness, or sampling variability, or noise. Correctly judging whether \emph{no} pattern exists in a residual plot is a difficult task. \cite{loy2021bringing} emphasizes that this is especially difficult to teach to new analysts and students, and advocates to the broader use of the lineup protocol \citep{bujastatistical2009}.}

\textcolor{red}{The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is \emph{no} pattern. (Figure \ref{fig:first-example-lineup} shows an example of a lineup of residual plots.) In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer doesn't know which is the true plot. When used this way, it provides an objective test for data plots. \cite{majumdervalidation2013} validated that results from lineups assessed by human observers performed similarly to conventional tests. One wouldn't use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by \cite{loy2013diagnostic}. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.}

@@

@@

During the experiment, every participant recruited from the Prolific crowd-sourcing platform [@palan2018prolific] is presented with a block of 20 lineups. A lineup consists of a randomly placed data plot and 19 null plots, which are all residual plots drawn with raw residuals on the y-axis and fitted values on the x-axis. An additional horizontal red line is added at $y = 0$ as a visual reference. The data in the data plot is simulated from one of two models described in Section \ref{simulating-departures-from-good-residuals}, while the data of the remaining 19 null plots are generated by the residual rotation technique discussed in Section \ref{visual-test-procedure-based-on-lineups}.

@@

Overall, we collected `r vi_survey  %>% filter(!attention_check) %>% nrow()` evaluations on `r vi_survey  %>% filter(!attention_check) %>% pull(unique_lineup_id) %>% unique() %>% length()` unique lineups performed by `r vi_survey  %>% filter(!attention_check) %>% count(exp, set) %>% nrow()` participants throughout three data collection periods.

A summary of the factors used in the experiment can be found in Table \ref{tab:model-factor-table}. There were four levels of the non-linear structure, and three levels of heteroskedastic structure. The signal strength was controlled by error variance ($\sigma$) for the non-linear pattern, and by a ratio ($b$) parameter for the heteroskedasticity. Additionally, three levels of sample size ($n$) and four different fitted value distributions were incorporated.

@@

Data collection period I was designed to study the ability of participants to detect the effect of a non-linear term $\boldsymbol{z}$ constructed using Hermite polynomials on random vector $\boldsymbol{x}$ formulated as

$$ \boldsymbol{y} = \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{z} + \boldsymbol{\varepsilon},~\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n)$$

\noindent where $\boldsymbol{y}$, $\boldsymbol{x}$, $\boldsymbol{\varepsilon}$, $\boldsymbol{z}$ are vectors of size $n$, $\boldsymbol{1}_n$ is a vector of ones of size $n$, $\boldsymbol{z} \propto He_j(\boldsymbol{x})$, $He_{j}(.)$ is the $j$th-order probabilist's Hermite polynomials [@hermite1864nouveau; originally by @de1820theorie].

To fit a misspecified model, the $\boldsymbol{z}$ term, which is a higher order term of $\boldsymbol{x}$, was omitted from the null regression formula

\begin{equation} \label{eq:null-model}
\boldsymbol{y} = \beta_0\boldsymbol{1}_n + \beta_1 \boldsymbol{x} + \boldsymbol{u},~ \boldsymbol{u} \sim N(\boldsymbol{0}_n, \sigma^2\boldsymbol{I}_n).
\end{equation}

Four different values of $j = 2, 3, 6, 18$ were chosen so that distinct shapes of non-linearity were included in the residual plot. These include "U", "S", "M" and "triple-U" shape as shown in Figure \ref{fig:different-shape-of-herimite}. A greater value of $j$ will result in a curve with more turning points. It is expected that the "U" shape will be the easiest to detect, and as the shape gets more complex it will be harder to perceive in a scatterplot, particularly when there is noise. Figure \ref{fig:different-sigma} shows the "U" shape for different amounts of noise ($\sigma$).

Data collection period II was similar to Data collection I but focuses on heteroskedasticity visual patterns formulated as

$$\boldsymbol{y} = \boldsymbol{1}_n + \boldsymbol{x} + \boldsymbol{\varepsilon},~\boldsymbol{\varepsilon} \sim N(\boldsymbol{0}_n, \boldsymbol{1}_n + (2 - |a|)(\boldsymbol{x} - a\boldsymbol{1}_n)'(\boldsymbol{x} - a\boldsymbol{1}_n) b \boldsymbol{I}_n),$$
\noindent where $\boldsymbol{y}$, $\boldsymbol{x}$, $\boldsymbol{\varepsilon}$ are vectors of size $n$.

The null regression model used to fit the realizations generated by the above model is formulated exactly the same as Equation \ref{eq:null-model}. For $b \neq 0$, the variance-covariance matrix of the error term $\boldsymbol{\varepsilon}$ is correlated with the predictor $\boldsymbol{x}$, which will lead to the presence of heteroskedasticity.

Visual patterns of heteroskedasticity are simulated using three different shapes ($a$ = -1, 0, 1) including "left-triangle", "butterfly" and "right-triangle" shapes as displayed in Figure \ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows the butterfly shape as the ratio parameter ($b$) is changed.

More details about the design of the data generating process are provided in the Appendix.

@@

Fitted values are a function of the independent variables, and the distribution of the observed values affects the distribution of the fitted values. In the best case scenario the fitted values will have a uniform distribution, which means that there is even coverage of possible observed values across all of the predictors. This is not always present in the collected data. Sometimes the fitted values are discrete because one or more predictors were measured discretely.  It is also common to see a skewed distribution of fitted values if one or more of the predictors has a skewed distribution. This latter problem is usually corrected before modelling, using a variable transformation. Our simulation assess this by using four different distributions to represent fitted values, constructed by different sampling of the raw predictor $X_{raw}$, namely, uniform ($U(-1, 1)$), normal ($N(0, 0.3^2)$), skewed ($\text{lognormal}(0, 0.6^2)/3$) and discrete ($U\{1, 5\}$).

@@

Three different sample sizes were used in our experiment: $n = 50, 100, 300$. Figure \ref{fig:different-n} shows the non-linear "S" shape for different sample sizes. We expect signal strength to decline in the simulated data plots with smaller $n$. We chose 300 as the upper limit, because it is typically enough for structure to be visible in a scatterplot reliably. \textcolor{red}{Beyond 300, the scatterplot should probably be used with transparency or replaced with a density or binned plot as scatterplots suffer from over-plotting.}

@@

For the non-linearity model, the key items defining effect size are sample size ($n$) and variance of the error term ($\sigma^2$), and so effect size would be roughly calculated as $\sqrt{n}/{\sigma}$. As sample size increases the effect size would increase, but as variance increases the effect size decreases. However, it is not clear how the additional parameter for the model polynomial order, $j$, should be incorporated. Intuitively, the large $j$ means more complex pattern, which likely means effect size would decrease. For the purposes of our calculations we have chosen to use an approach based on Kullback-Leibler divergence [@kullback1951information], coupled with simulation. This formulation defines effect size to be:

\begin{equation}
E = \frac{1}{2}\left(\boldsymbol{\mu}_z'(\text{diag}(\boldsymbol{R}\sigma^2))^{-1}\boldsymbol{\mu}_z\right)
\label{eq:nonlin-effect-size}
\end{equation}

\noindent where $\text{diag}(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{H}$ is the residual operator, $\boldsymbol{H} = \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the hat matrix, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contain any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}_n$ is the assumed covariance matrix of the error term when $H_0$ is true.

In the heteroskedasticity model, the key elements for measuring effect size are sample size, $n$, and the ratio of the biggest variance to smallest variance, $b$. Larger values of both would produce higher effect size. However, it is not clear how to incorporate the additional shape parameter, $a$. Thus the same approach is used here, where the formula can be written as:

\begin{equation}
E = \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|\text{diag}(\boldsymbol{R})|} - n + \text{tr}(\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\text{diag}(\boldsymbol{R}))\right)
\label{eq:hetero-effect-size}
\end{equation}

\noindent where $\boldsymbol{V}$ is the actual covariance matrix of the error term. Derivations for these equations are provided in the Appendix.

@@

# \textcolor{red}{Limitations and practicality}

@@

These `R` packages are used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor]. The study website is powered by `PythonAnywhere` [@pythonanywhere] and the `Flask` web framework [@flask]. The `jsPsych` framework [@jspsych] is used to create behavioural experiments that run in our study website.

>>>>

Diagnosing a model is an important part of building an appropriate model. In linear regression analysis, studying the residuals from a model fit is a common diagnostic activity. Residuals summarise what is not captured by the model, and thus provide the capacity to identify what might be wrong.

@@

Scatterplots of residuals against the fitted values, and each of the explanatory variables, are commonly used to scrutinize their relationships. If there are any visually discoverable associations, the model is potentially inadequate or incorrectly specified. We can also potentially discover patterns not directly connected to a linear model assumption from these residual plots, such as the discreteness or skewness of the fitted values, and outliers. To read residual plots, one looks for noticeable departures from the model such as non-linear pattern or heteroskedasticity. A non-linear pattern would suggest that the model needs to have some additional non-linear terms. Heteroskedasticity suggests that the error is dependent on the predictors, and hence violates the independence assumption. Statistical tests were developed to provide objective assessment, for example, of non-linear patterns [e.g. @ramsey1969tests], and heteroskedasticity  [e.g. @breusch1979simple].

@@

Additionally, relying solely on the subjective assessment of a single plot can be problematic. People will almost always see a pattern \citep[see][]{kahneman2011thinking}, so the question that really needs answering is whether any pattern perceived is consistent with randomness, or sampling variability, or noise. Correctly judging whether \emph{no} pattern exists in a residual plot is a difficult task. \cite{loy2021bringing} emphasizes that this is especially difficult to teach to new analysts and students, and advocates to the broader use of the lineup protocol \citep{bujastatistical2009}.

The lineup protocol places a data plot in a field of null plots, allowing for a comparison of patterns due purely by chance to what is perceived in the data plot. For residual analysis this is especially helpful for gauging whether there is \emph{no} pattern. (Figure \ref{fig:first-example-lineup} shows an example of a lineup of residual plots.) In its strict use, one would insist that the data plot is not seen before seeing the lineup, so that the observer doesn't know which is the true plot. When used this way, it provides an objective test for data plots. \cite{majumdervalidation2013} validated that results from lineups assessed by human observers performed similarly to conventional tests. One wouldn't use a lineup when a conventional test exists and is adequate because it is more manually expensive to conduct. However, where no adequate conventional test exists, it is invaluable, as shown by \cite{loy2013diagnostic}. Here we use the lineup as a vehicle to rigorously explore why experts advise that residual plots are indispensable despite the prevalence of numerical tests.

@@

\vspace{-\baselineskip}

@@

During the experiment, every participant recruited from the Prolific crowd-sourcing platform [@palan2018prolific] was presented with a block of 20 lineups. A lineup consisted of a randomly placed data plot and 19 null plots, which were all residual plots drawn with raw residuals on the y-axis and fitted values on the x-axis. An additional horizontal red line was added at $y = 0$ as a visual reference. The data in the data plot was simulated from one of two models described in Section \ref{simulating-departures-from-good-residuals}, while the data of the remaining 19 null plots were generated by the residual rotation technique discussed in Section \ref{visual-test-procedure-based-on-lineups}.

@@

Overall, we collected `r vi_survey  %>% filter(!attention_check) %>% nrow()` evaluations on `r vi_survey  %>% filter(!attention_check) %>% pull(unique_lineup_id) %>% unique() %>% length()` unique lineups performed by `r vi_survey  %>% filter(!attention_check) %>% count(exp, set) %>% nrow()` participants. A summary of the factors used in the experiment can be found in Table \ref{tab:model-factor-table}. There were four levels of the non-linear structure, and three levels of heteroskedastic structure. The signal strength was controlled by error variance ($\sigma$) for the non-linear pattern, and by a ratio ($b$) parameter for the heteroskedasticity. Additionally, three levels of sample size ($n$) and four different fitted value distributions were incorporated.

@@

Data collection period I was designed to study the ability of participants to detect non-linearity departures from residual plots. The non-linearity departure was constructed by omitting a $j$th order Hermite polynomial [@hermite1864nouveau; originally by @de1820theorie] term of the predictor from the simple linear regression equation. Four different values of $j = 2, 3, 6, 18$ were chosen so that distinct shapes of non-linearity were included in the residual plots. These include "U", "S", "M" and "triple-U" shape as shown in Figure \ref{fig:different-shape-of-herimite}. A greater value of $j$ will result in a curve with more turning points. It is expected that the "U" shape will be the easiest to detect, and as the shape gets more complex it will be harder to perceive in a scatterplot, particularly when there is noise. Figure \ref{fig:different-sigma} shows the "U" shape for different amounts of noise ($\sigma$).

Data collection period II was similar to period I but focuses on heteroskedasticity departures. We generated the heteroskedasticity departures by setting the variance-covariance matrix of the error term as a function of the predictor, but fitted the data with the simple linear regression model, intentionally violated the constant variance assumption. Visual patterns of heteroskedasticity are simulated using three different shapes ($a$ = -1, 0, 1) including "left-triangle", "butterfly" and "right-triangle" shapes as displayed in Figure \ref{fig:different-shape-of-heter}. Figure \ref{fig:different-b} shows the butterfly shape as the ratio parameter ($b$) is changed. More details about the simulation process are provided in the Appendix.

@@

Fitted values are a function of the independent variables, and the distribution of the observed values affects the distribution of the fitted values. In the best case scenario the fitted values will have a uniform distribution, which means that there is even coverage of possible observed values across all of the predictors. This is not always present in the collected data. Sometimes the fitted values are discrete because one or more predictors were measured discretely.  It is also common to see a skewed distribution of fitted values if one or more of the predictors has a skewed distribution. This latter problem is usually corrected before modelling, using a variable transformation. Our simulation assess this by using four different distributions to represent fitted values, constructed by different sampling of the predictor, including $U(-1, 1)$ (uniform), $N(0, 0.3^2)$ (normal), $\text{lognormal}(0, 0.6^2)/3$ (skewed) and $U\{-1, 1\}$ (discrete).

@@

Three different sample sizes were used in our experiment: $n = 50, 100, 300$. Figure \ref{fig:different-n} shows the non-linear "S" shape for different sample sizes. We expect signal strength to decline in the simulated data plots with smaller $n$. We chose 300 as the upper limit, because it is typically enough for structure to be visible in a scatterplot reliably. Beyond 300, the scatterplot should probably be used with transparency or replaced with a density or binned plot as scatterplots suffer from over-plotting.

@@

For the non-linearity model, the key items defining effect size are sample size ($n$) and the noise level ($\sigma^2$), and so effect size would be roughly calculated as $\sqrt{n}/{\sigma}$. Increasing sample size tends to boost the effect size, while heightened noise diminishes it. However, it is not clear how the additional parameter for the model polynomial order, $j$, should be incorporated. Intuitively, the large $j$ means more complex pattern, which likely means effect size would decrease. Similarly, in the heteroskedasticity model, effect size relies on sample size ($n$) and the ratio of the largest to smallest variance, $b$. Larger values of both would produce higher effect size, but the role of the additional shape parameter, $a$, in this context is unclear.

For the purposes of our calculations we have chosen to use an approach based on Kullback-Leibler divergence [@kullback1951information], coupled with simulation. This formulation defines effect size to be

\vspace{-\baselineskip}

$$E = \frac{1}{2}\left(\log\frac{|\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')|}{|\text{diag}(\boldsymbol{R}\sigma^2)|} - n + \text{tr}(\text{diag}(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\text{diag}(\boldsymbol{R}\sigma^2)) + \boldsymbol{\mu}_z'(\boldsymbol{R}\boldsymbol{V}\boldsymbol{R}')^{-1}\boldsymbol{\mu}_z\right),$$

\noindent where $\text{diag}(.)$ is the diagonal matrix constructed from the diagonal elements of a matrix, $\boldsymbol{X}$ is the design matrix, $\boldsymbol{V}$ is the actual covariance matrix of the error term,
$\boldsymbol{R} = \boldsymbol{I}_n - \boldsymbol{X}(\boldsymbol{X}'\boldsymbol{X})^{-1}\boldsymbol{X}'$ is the residual operator, $\boldsymbol{\mu}_z = \boldsymbol{R}\boldsymbol{Z}\boldsymbol{\beta}_z$ is the expected values of residuals where $\boldsymbol{Z}$ contains any higher order terms of $\boldsymbol{X}$ left out of the regression equation, $\boldsymbol{\beta}_z$ contains the corresponding coefficients, and $\sigma^2\boldsymbol{I}_n$ is the assumed covariance matrix of the error term when $H_0$ is true.

@@

# Limitations and practicality

@@

These `R` packages were used for the work: `tidyverse` [@tidyverse], `lmtest` [@lmtest], `mpoly` [@mpoly], `ggmosaic` [@ggmosaic], `kableExtra` [@kableextra], `patchwork` [@patchwork], `rcartocolor` [@rcartocolor]. The study website was powered by `PythonAnywhere` [@pythonanywhere] and `Flask` web framework [@flask]. The `jsPsych` framework [@jspsych] was used to create behavioural experiments that run in our study website.
